---
title: 'Project Presentation Rubrics'
author: 'STA321 - Advance Topics in Statistics'
date: ""
output:
  html_document: 
    toc: no
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    fig_caption: yes
    number_sections: yes
    theme: readable
  pdf_document: 
    toc: no
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 5
    fig_height: 4
  word_document: 
    toc: no
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
---
<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0;
}
h1.title {
  font-size: 20px;
  color: DarkRed;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}
h1 { /* Header 3 - and the author and data headers use this too  */
    font-size: 22px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}
h2 { /* Header 3 - and the author and data headers use this too  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 15px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}
</style>

```{r setup, include=FALSE}
# install.packages("remotes") through github since it not
# available from CRAN. 
# 
# remotes::install_github("rlesur/klippy")
##
# install.packages("webshot")
# webshot::install_phantomjs()

# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
library(knitr)
library(klippy)
library(webshot)
library(MASS)
knitr::opts_chunk$set(echo = TRUE,      # include code chunk in the output file
                      warnings = FALSE, # sometimes, you code may produce warning messages,
                                        # you can choose to include the warning messages in
                                        # the output file. 
                      results = TRUE    # you can also decide whether to include the output
                                        # in the output file.
                      )  
```

```{r klippy, echo=FALSE, include=TRUE}
#klippy::klippy(position = c('top', 'right'),color = 'darkred',tooltip_message = 'Click to copy', tooltip_success = 'Done')
```



\

The total possible points is 10. The weights of all components are given in the following table of the common rubrics for all weekly assignments.

\

|Rubric for Weekly Assignments                          |Points Possible|Points Earned|
|-------------------------------------------------------|---------------|-------------|
|**Introduction/Title**:	                              |1.5	          |             |
|Title is clear and reflects the analysis             	|0.5	          |             |
|Clearly describes the question that is being addressed	|0.5            |             |
|Clearly states the research question of interest	      |0.5            |	            |
|                                                       |               |             |
|**Data Collection**:	                                      |1            	|             |    
|The method of data collection is clearly described	|0.5|	|
|List of the variables and their corresponding definitions|	0.5||	
|                                                       |               |             |		
|**Exploratory Data Analysis**:	|1.5||	
|Variable inspection, transformation, and standardization	|0.5|	|
|Pairwise scatter plot and potential collinearity	|0.5|	|
|Summary statistics		|0.5| |
|                                                       |               |             |
|**Modeling and Inferential Statistics**:	|2.5	||
|Appropriate models	|0.5|	|
|Model search process	|1|	|
|Final model and justification|	0.5	||
|Model diagnostics and interpretation	|0.5	||
|                                                       |               |             |
|**Discussion and Conclusions**:|	2	||
|Clearly and correctly addresses the question of interest	|0.5	||
|Concise, accurate, and straightforward summary of the findings|	0.5|	|
|Drawbacks and limitations of the analysis|	0.5	||
|Improvements and generalizations	|0.5	||
|                                                       |               |             |
|**Overall Impression**:	|1.5	||
|Figures and tables included in the report must support some arguments	|0.5||	
|Organized to answer the question of interest	|0.5||	
|Question of interest is non-trivial and well-formulated	|0.5	||


\

\

\







